### 2009-12-07 docs/marcello-dissertation/thesis/chapter10-appendix.tex
Most of the revisions and updates... Changes to chapters 3, 4 after Kleber's chat online (adding missing taxonomy).
Revisions of print chapters until chapter 5.

Chapter 7 is reviewed until before discussion. 

Next revision should include all the necessary reviews for the
release draft.

git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@676 559edde2-a4d4-44e0-96fc-4035e40d7396

\section{mongoDB Deployment}

@@

\subsection{Starting the Database}

\subsection{Stopping the Database}

\section{Experiment Execution}

@@

\subsection{Data Access Using Database Shell}

@@

\lstset{label=cmd:mongo-rest-request,caption=REST HTTP GET Request Example}
\begin{lstlisting}
HTTP/1.0 200 OK
x-action:
x-ns: netbeams.SondeDataContainer
Content-Type: text/plain;charset=utf-8

{
  "offset" : 0,
  "rows": [
    { "_id" : "156f4007e4c3b74a36ed3100", "sensor_ip_address" : "192.168.0.117", "message_id" : "08b02c08-9290-4517-9a28-c6ee7e16509a",
"transaction_time" : 1253557219486, "fact_time" : 1253557217000, "data" : { "temperature" : "31.44", "sp_condition" : "99.8", "condition" : "53.5",
"resistence" : "1157.08", "salinity" : "0.0", "pressure" : "1.066", "depth" : "0.161", "ph" : "1.08", "pH_mv" : "-82.0", "odo_sat" : "40.3",
"odo_condition" : "56.85", "turbidity" : "0.2", "battery" : "8.2" } } ,
    { "_id" : "156f4007e5c3b74a37ed3100", "sensor_ip_address" : "192.168.0.117", "message_id" : "08b02c08-9290-4517-9a28-c6ee7e16509a",
"transaction_time" : 1253557219486, "fact_time" : 1253557217000, "data" : { "temperature" : "37.83", "sp_condition" : "176.3", "condition" : "2.6",
"resistence" : "1324.97", "salinity" : "0.01", "pressure" : "1.36", "depth" : "1.564", "ph" : "0.12", "pH_mv" : "-23.5", "odo_sat" : "104.5",
"odo_condition" : "19.44", "turbidity" : "0.1", "battery" : "5.0" } } ,
    { "_id" : "156f4007e5c3b74a38ed3100", "sensor_ip_address" : "192.168.0.117", "message_id" : "08b02c08-9290-4517-9a28-c6ee7e16509a",
"transaction_time" : 1253557219486, "fact_time" : 1253557217000, "data" : { "temperature" : "74.3", "sp_condition" : "84.0", "condition" : "104.7",
"resistence" : "4089.13", "salinity" : "0.01", "pressure" : "1.222", "depth" : "2.788", "ph" : "6.56", "pH_mv" : "-78.1", "odo_sat" : "40.0",
"odo_condition" : "6.02", "turbidity" : "0.3", "battery" : "3.2" } },
  "total_rows" : 3 ,
  "query" : {} ,
  "millis" : 0
}
\end{lstlisting}

@@

An example to export the data in CSV format can be seen in listing
\ref{cmd:mongoexport}, and an example containing 1 million objects can be
downloaded at
http://netbeams.googlecode.com/files/experiment-1000000-data-exported-20090913-053538.csv.tar.gz.

\lstset{label=cmd:mongoexport,caption=Command to export data in CSV format}
\begin{lstlisting}
mongoexport -d netbeams -c SondeDataContainer --dbpath ./data/ --csv -f "_id,sensor_ip_address,transaction_time,fact_time,
data.temperature,data.sp_condition,data.condition,data.resistence,data.salinitude,data.pressure,data.depth,data.ph,data.pH_mv,data.odo_sat,
data.odo_condition,data.turbidity,data.battery" -o sonde-data-exported.csv
\end{lstlisting}

@@

\lstinputlisting[language=Bash,label=file:experiment-query-scenarios,caption=Experiment
Scenarios Implementation]{../../../../netbeams/versions/v2/persistence/experiment-client-scenarios.js}

@@

@@

Sun Nov 29 19:07:04 query netbeams.SondeDataContainer ntoreturn:0 reslen:2258 nscanned:101 { query: {}, $snapshot: 1 }  nreturned:101 202ms
Sun Nov 29 19:07:04 getmore netbeams.SondeDataContainer cid:7749363894863150628 ntoreturn:0 query: { query: {}, $snapshot: 1 }  bytes:1048562 nreturned:47661 1767ms
...
...
...
Sun Nov 29 19:10:18 getmore netbeams.SondeDataContainer cid:7749363894863150628 ntoreturn:0 query: { query: {}, $snapshot: 1 }  bytes:793098 nreturned:36049 1975ms
exported 2419200 records
\end{lstlisting}

>>>>

\section{mongoDB Deployment}
\label{sec:mongodb-deployment}

@@

\subsection{MongoDB Database Administration}

- starting the db
- Stopping the Database

\section{Experiment Execution}
\label{sec:experiment-execution}

@@

\subsection{Data Access Using Database Shell}
\label{sec:mongodb-user-experience}

@@

\lstset{label=cmd:mongo-rest-request,caption=REST HTTP GET Request Example}
\begin{lstlisting}
http://192.168.1.2:28017/netbeams/SondeDataContainer/?limit=-3&filter_observation.Conductivity=104.5

HTTP/1.0 200 OK
x-action:
x-ns: netbeams.SondeDataContainer
Content-Type: text/plain;charset=utf-8

{
  "offset" : 0,
  "rows": [
    { "_id" : { "$oid" : "616f400784ba194bc2dee200" }, "message_id" : "c76a9898-1eee-4463-bf5d-d40a7a8af6e3", "sensor" : { "ip_address" : "192.168.0.60", "location" : { "latitude" : 37.89155, "longitude" : -122.4464 } }, "time" : { "valid" : { "$date" : 1261229067000 }, "transaction" : { "$date" : 1259977271704 } }, "observation" : { "WaterTemperature" : 51.29, "SpecificConductivity" : 110.7, "Conductivity" : 104.5, "Resistivity" : 1710.07, "Salinity" : 0.01, "Pressure" : 1.094, "Depth" : 1.808, "pH" : 3.04, "pHmV" : -46.4, "Turbidity" : 0.2, "ODOSaturation" : 59, "ODO" : 10.66, "Battery" : 7.1 } } ,
    { "_id" : { "$oid" : "616f400789ba194bdbe2e200" }, "message_id" : "2eb3f5ca-7a30-4b5d-8d16-c3e80a7b1b7a", "sensor" : { "ip_address" : "192.168.0.24", "location" : { "latitude" : 37.89155, "longitude" : -122.4464 } }, "time" : { "valid" : { "$date" : 1261740924000 }, "transaction" : { "$date" : 1259977271704 } }, "observation" : { "WaterTemperature" : 10.98, "SpecificConductivity" : 100.9, "Conductivity" : 104.5, "Resistivity" : 5448.21, "Salinity" : 0.08, "Pressure" : 1.152, "Depth" : 1.289, "pH" : 3.09, "pHmV" : -0.4, "Turbidity" : 0.1, "ODOSaturation" : 72.2, "ODO" : 59.65, "Battery" : 6.3 } } ,
    { "_id" : { "$oid" : "616f400789ba194b13e5e200" }, "message_id" : "a7804d0a-44a8-4bd9-a9c7-51db84e91f08", "sensor" : { "ip_address" : "192.168.0.236", "location" : { "latitude" : 37.89155, "longitude" : -122.4464 } }, "time" : { "valid" : { "$date" : 1260969259000 }, "transaction" : { "$date" : 1259977271704 } }, "observation" : { "WaterTemperature" : 7.97, "SpecificConductivity" : 119.1, "Conductivity" : 104.5, "Resistivity" : 1971.59, "Salinity" : 0.05, "Pressure" : 1.516, "Depth" : 2.187, "pH" : 3.11, "pHmV" : -77.59999999999999, "Turbidity" : 0.2, "ODOSaturation" : 82.90000000000001, "ODO" : 41.99, "Battery" : 0.1 } }
  "total_rows" : 3 ,
  "query" : {} ,
  "millis" : 0
}
\end{lstlisting}

@@

An example to export the data in CSV format can be seen in listing
\ref{file:mongodb-export-command}, and an example containing 1 million objects can be
downloaded at
http://netbeams.googlecode.com/files/experiment-1000000-data-exported-20090913-053538.csv.tar.gz.

@@

\lstinputlisting[language=Python,label=file:experiment-setup-mongoshards,caption=Experiment
Main Execution Script]{../../../../netbeams/versions/v2/persistence/setup-mongo-shards.js}

\lstinputlisting[language=Python,label=file:experiment-export-python,caption=Data
Export to from mongoDB to OPEnDAP format.]{../../../../netbeams/versions/v2/persistence/data-export-python.py}

\lstinputlisting[language=Python,label=file:experiment-scenarios,caption=Experiment
Scenarios Implementation]{../../../../netbeams/versions/v2/persistence/experiment-client-scenarios.js}

\lstinputlisting[language=Java,label=file:main-ysi-data-handler,caption=The
DSP YSI Sonde Data Handler]{../../../../netbeams/versions/v2/apps/xml/jaxb/sonde/java-src/org/netbeams/dsp/ysi/SondeDataType.java}

@@

\subsection{Experiment Execution Output}

@@

Sun Nov 29 19:07:04 query netbeams.SondeDataContainer ntoreturn:0 reslen:2258 nscanned:101 { query: {}, $snapshot: 1 }  nreturned:101 202ms
Sun Nov 29 19:07:04 getmore netbeams.SondeDataContainer cid:7749363894863150628 ntoreturn:0 query: { query: {}, $snapshot: 1 }  bytes:1048562 nreturned:47661 1767ms
...
...
Sun Nov 29 19:10:18 getmore netbeams.SondeDataContainer cid:7749363894863150628 ntoreturn:0 query: { query: {}, $snapshot: 1 }  bytes:793098 nreturned:36049 1975ms
exported 2419200 records
\end{lstlisting}

\lstinputlisting[label=file:experiment-export-python-results,caption=Data
Export to from mongoDB to OPEnDAP format.]{../../../../netbeams/versions/v2/persistence/export-mongodb-to-sfbeams.csv}

### 2009-12-07 docs/marcello-dissertation/thesis/chapter2-state-of-the-art.tex
Most of the revisions and updates... Changes to chapters 3, 4 after Kleber's chat online (adding missing taxonomy).
Revisions of print chapters until chapter 5.

Chapter 7 is reviewed until before discussion. 

Next revision should include all the necessary reviews for the
release draft.

git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@676 559edde2-a4d4-44e0-96fc-4035e40d7396

@@

\section{Data Provenance: How Collected Data is Described?}
\label{sec:sn-provenance}

As described in the previous section, the collected data from sensor networks
can be represented by different data models, requiring the knowledge of the
properties of the data. Not to mention that the low-level or raw data produced
by a given sensor device might not include vital information regarding the
origing and identity of the collected data. For this reason,
\cite{sn-provenance} describes Data Provenance as a technique used to describe
those essential properties of the collected data from sensors. One typical
example of the application of Data Provenance is the so-called blackbox of an
aircraft, which carries the collection of raw data collected from a group of
sensors placed in different locations of an aircraft, as shown by
\cite{sn-exemple-blackbox}. Last, but not least, the author describes the
collected data as used in a process of data recovery and analysis in order to
identify the probable reason of an aircraft accident, by using the time, and
location of the aircraft, along with the independent variables from the
sensors.

@@

@@

In order to use the collected data for the purpose of historical analysis, the
use of a point in time is the most prominent feature of data, as it is
considered a pivotal metadata for different types of sensor networks such as
an aircraft the blackbox \cite{sn-exemple-blackbox} or any
environmental sensor network \cite{sfbeams2006, sn-xml-middleware, sn-ex01}.
In this way, Other projects classify data in time-series
\cite{sn-time-series-example}, where data is related to all the
observations a given sensor device performed over a period of time. The
apparent implications is that all the produced data is necessary, but
\cite{sn-time-series} argued that some sorts of sensor networks could decrease
the creation of ``noisy data sets" based on the proper clustering of
similar time-series data before before they are sent to the data sink.

Based on the guidelines defined by \cite{db-temporal}, the time properties of
the collected data can be divided into two different dimensions: \textbf{valid
time} and the \textbf{transaction time}. The former depicts the point in time
when the observation took place with respect to the real-world space. An
important aspect of this metadata is that since it captures the time when the
observation occurred, its value cannot be changed during its existence, as well
as placing restrictions on the update operations of this constant. Similarly,
the latter type of is the point in time at which a given fact was transferred
to the persistence layer.

@@

>>>>

It is important to note that the scope of this literature review is limited to
the scope of the our case study described in chapter 4 and, therefore, it only
covers a subset of disciplines in sensor networks and data persistence.

@@

\section{How collected data is described?}
\label{sec:sn-data-description}

As described in the previous section, the collected data from sensor networks
can be represented by different data models, requiring the knowledge of the
properties of the data. Not to mention that the low-level or raw data produced
by a given sensor device might not include vital details regarding the
origing and identity of the collected data. For this reason, the collected data
is reported to be described in different ways such as Data Provenance or
Annotations.

\cite{sn-provenance} describes Data Provenance as a technique used to describe
those essential properties of the collected data from sensors. One typical
example of the application of Data Provenance is the so-called blackbox of an
aircraft, which carries the collection of spacio-temporal data structures
collected from a group of sensors placed in different locations of an
aircraft, as shown by \cite{sn-exemple-blackbox}. Last, but not least, the
author describes the collected data as used in a process of data recovery and
analysis in order to identify the probable reason of an aircraft accident, by
using the time, and location of the aircraft, along with the independent
variables from the sensors.

@@

Another different way to describe sensor devices is through the use of
Annotations. \cite{sn-annotation} shows the use of tags to describe data
collected from a physical area monitoring. The sensor network was
designed to monitor a moving objects, capturing their location while in
movement and their visual appearance using a video camera. In this way,
different instants can be described by adding tags particular video frame, and
therefore, providing an efficient way to search of a particular frame based on
its visual observations. Similarly, this project also uses data provenance to
describe its collected data, since each of the collected data contains the
related GPS coordinates of the observed object.

@@

In order to use the collected data for the purpose of historical analysis, the
use of a point in time is the most prominent feature of data. Also referred to
as spatial-temporal data, it is considered a pivotal metadata used in different
types of sensor networks such as an aircraft the blackbox
\cite{sn-exemple-blackbox}, environmental sensor network \cite{sfbeams2006},
and data communications management systems \cite{sn-dataware-house}. In this
way, Other projects classify data in time-series \cite{sn-time-series-example},
where data is related to all the observations a given sensor device performed
over a period of time. The apparent implications is that all the produced data
is necessary, but \cite{sn-time-series} argued that some sorts of sensor
networks could decrease the creation of ``noisy data sets" based on the proper
clustering of similar time-series data before before they are sent to the data
sink.

Based on the guidelines defined by \cite{db-temporal}, the time properties of
the collected data can be divided into two different dimensions: \textbf{valid
time} and the \textbf{transaction time}. The former depicts the point in time
when the observation took place with respect to the real-world space. An
important aspect of this metadata is that since it captures the time when the
observation occurred, its value cannot be changed during its existence, as well
as placing restrictions on the update operations of this constant. Similarly,
the latter type of is the point in time at which a given fact was transferred
to the persistence layer. Either types of time can be used depending on the
requirements of the sensor network. For instance, \cite{sn-dataware-house}
reports that all the data stored by its sensor network uses the time of
collection, but do not use the transaction time for that purpose.

@@

\section{Data Load and Management Systems}
\label{sec:data-load}

Depending on the amount of data produced by a sensor network, different
approaches are reported to be taken to support the data management depending on
the amount of data collected. For example, \cite{sfbeams2006} is a an
environmental sensor network that produces data regarding the marine life.
Less than twenty sensor devices produce Megabytes of data in a daily bases,
whose data is processed and stored in using flat-files directly in the
operating system's file system with provenance information regarding the time
of collection.

When the amount of data collected grows to other scales, say, Gigabyes of data
in a daily basis, data management tends to become more difficult in any type of
database systems. Examples of sensor networks that produce data under this
magnitude of data are reported in the literature. For instance, 
\cite{sn-dataware-house} describes the creation of a Data Warehouse for the
real-time sensor data management using a cluster of MySQL databases used to
store data from different sensors such as optical sensors that collect metrics
from physics experiments using eletromagnetic fields. As the author reported,
requirements for data management evolved from storing flat files by a
monolitic system to a robust distributed system using MySQL database.
Moreover, the the system targed from 5 to 10MB of data per second, whose
culmulative data reached 1 TB of data every year.

Data management for large scales of data require different techniques to store
and retrieve data, since it requires more computational power and data
retrieval efficient algorithms. \cite{sn-data-center} describes the creation of
a Sensor Data Center designed to store and analyze amounts of data on the range of
Terabytes to a future Petabyes of data. Its sensor network deals with not only
numerical sensor readings from transportation managemenet, but also radar
images, used for historic purposes and data mining that performs traffic
pattern analysis. In this way, the authors report the use of parallel data
processing techniques in order to provide processing for the collected
data, and load-balancing when storing data to minimize the sensor data
processing delays. For this reason, they suggest the use of concurrent control
such as snapshot isolation of the current data into separated replicated data.
In this way, data access from users are not interrupted by the processing of
the data that is processed.

### 2009-12-07 docs/marcello-dissertation/thesis/chapter3-taxonomic-classification-of-persistence-storage-for-sensor-networks.tex
Most of the revisions and updates... Changes to chapters 3, 4 after Kleber's chat online (adding missing taxonomy).
Revisions of print chapters until chapter 5.

Chapter 7 is reviewed until before discussion. 

Next revision should include all the necessary reviews for the
release draft.

git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@676 559edde2-a4d4-44e0-96fc-4035e40d7396

\begin{itemize}
  \item \textbf{Real-time Data Stream}: the collected data is accessed
  on-the-fly from the sensor device as a real-time data stream, and may have a
  short life cycle as it temporarily resides in memory;
  \item \textbf{Data Archival}: the collected data is used for historical
  purposes and is related to the historical data analysis \cite{sn-intro01,
  sn-intro02} or Information Fusion \ref{sn-info-fusion}. 
\end{itemize}

@@

\begin{itemize}
  \item \textbf{Local Storage}: characterizes the placed of the collected data
  temporarily located in-memory, or in a secondary storage device on a node
  located in-network;
  \item \textbf{External Storage}: The collected data is usually located in an
  external storage device with a data management system such as a relational
  database;
  \item \textbf{Data-Centric Storage}: when the collected data occupies
  different external storages, being organized by categories based on the
  collected data keys and values. In other words, what characterizes this taxon
  is the presence of a data partitioning strategy to store the collected data.
\end{itemize}

@@

\begin{figure}[h]
  \centering
  \includegraphics{../diagrams/taxonomy-data-model}
  \caption{Data Model Taxonomy}
  \label{fig:taxonomy-data-model}
\end{figure}

@@

\section{Data Provenance Taxonomy}

No matter which data model used to represent the collected data, Data
Provenance provides a valuable guideline on how to describe the collected data
from sensors devices. As reported in section \ref{sec:sn-provenance}, the
description of the properties should take into account different properties of
the nature of the data as shown in Figure \ref{fig:taxonomy-data-provenance}.
Those taxa are summarized as follows:

\begin{figure}[h]
  \centering
  \includegraphics{../diagrams/taxonomy-data-provenance}
  \caption{Data Provenance Taxonomy}
  \label{fig:taxonomy-data-provenance}
\end{figure}

\begin{itemize}
  \item \textbf{What: Data Identity}: the data type that uniquely identifies
  the collected data from sensors. First and foremost, the properties of the
  sensor devices must be described according to the meaningful naming
  convention for the property. For instance, ``Water Temperature'' better
  describes the intent of the property as ``temp''. Second, the identification
  of the collected data related to the sensor type should be given, that is,
  which sensor device created the collected data, in case the data origin is
  relevant;
  \item \textbf{When: Time Dimension}: time is an important attribute of the
  data because it identifies the age of the collected data. Depending on the 
  use of the collected data, this is often taken into account. Two different
  types of time dimensions were described: \textbf{fact} and
  \textbf{transaction} times. The former is used to identify when the data
  was collected, whereas the latter when the data was transmitted to the
  data sink where the query processing acts;
  \item \textbf{Where: Data Location}: the point where the collected data was
  observed by the sensor device. Usually the latitude and longitude as the GPS
  coordinates are provided by devices which are capable of sensing the
  location. A descriptive metadata about the location could also be given such
  as room-name = ``deep see''.
\end{itemize}

@@

\begin{figure}[h]
  \centering
  \includegraphics{../diagrams/taxonomy-query-mechanism}
  \caption{Query Processing Mechanism Taxonomy}
  \label{fig:taxonomy-query-mechanism}
\end{figure}

@@

\section{Database System Organization Taxonomy}

Different approaches of data systems were described in the literature review in
the last chapter in different sections. This taxonomy is directly related to
the use of external storage devices on a centralized network node so that it
often uses a single database server. Paradoxically, the use of a data-centric
storage determines the use of distributed database nodes around the network. 
In this way, Figure \ref{fig:taxonomy-database-architecture} describes the
taxonomy of the database system organization as follows:

\begin{figure}[h]
  \centering
  \includegraphics{../diagrams/taxonomy-database-architecture}
  \caption{Database Architecture Taxonomy}
  \label{fig:taxonomy-database-architecture}
\end{figure}

\begin{itemize}
  \item \textbf{Centralized System}: a database system that runs in one single
  centralized host, being the focal purpose of data traffic as read and writes 
  \cite{sn-intro01};
  \item \textbf{Distributed System}: a database system that can be composed by
  more than one node. Usually, the use of the pattern of Master-Slave
  mechanism or data ``Replication" is used. Another mechanism is the use of
  database paritioning, if the technology supports it.
\end{itemize}

@@

The different taxonomies defined in the previous sections of this chapter are
important to the understanding of a data persistence layer for sensor
networks. As they are closely related to each other, it is important to
briefly highlight these relationships in order to better analyze a problem
related to data persistence.

The taxonomy related to the purpose of the sensor data can be directly related
to the taxonomy of the location of the sensor data. While the taxon of the 
real-time data stream relates to the local data storage, the taxon related to
data archival correlates the storage of the collected data into an external
storage. In addition to that, the converse is also true.

The taxonomy related to the data model can be related to any taxon from the
taxonomy of the location of the sensor data, as well as the taxons related to
Data Provenance. In the former, the data of a relational model is stored using
SQL messages.

>>>>

\begin{itemize}
  \item \textbf{Real-time Data Stream}: the collected data is accessed
  on-the-fly from the sensor device as a real-time data stream, and may have a
  short life cycle as it temporarily resides in memory;
  \item \textbf{Data Archival}: the collected data is used for historical
  purposes and is related to the historical data analysis \cite{sn-intro01,
  sn-intro02} or Information Fusion \cite{sn-info-fusion}. 
\end{itemize}

@@

\begin{itemize}
  \item \textbf{Local Storage}: characterizes the placed of the collected data
  temporarily located in-memory, or in a secondary storage device on a node
  located in-network;
  \item \textbf{External Storage}: The collected data is usually located in an
  external storage device with a data management system such as a relational
  database. When the collected data occupies different external storages, the
  strategy is called \textbf{Data-Centric Storage}, being organized by
  categories based on the collected data keys and values. In other words, what
  characterizes this taxon is the presence of a data partitioning strategy to
  store the collected data.
\end{itemize}

@@

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.5]{../diagrams/taxonomy-data-model}
  \caption{The Data Model Taxonomy}
  \label{fig:taxonomy-data-model}
\end{figure}

@@

\section{Data Description Taxonomy}

Particulaly related to how the collected data is described, different projects
provide data persistence describing the collected data by using its natural
properties, or associating textual descriptions or keywords of the data. The
way to describe data can follow different guidelines such as data provenance
and annotations, discussed in section \ref{sec:sn-data-description}. As
reported in section \ref{sec:sn-data-description}, the description of the
properties should take into account different properties of the nature of the
data and any other description needed. Therefore, the taxa that composes the
Data Description taxonomy, shown in Figure
\ref{fig:taxonomy-data-description}, are described as follows:

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.5]{../diagrams/taxonomy-data-description}
  \caption{The Data Description Taxonomy}
  \label{fig:taxonomy-data-description}
\end{figure}

\begin{itemize}
  \item \textbf{Provenance-Aware}: when the collected data is categorized based
  on the nature of the data, \textbf{What: Data Identity} is the data type
  that uniquely identifies the collected data from sensors. Similarly, the
  \textbf{When: Time Dimension}, is related instants of time related to the
  collected data: \textbf{fact} and \textbf{transaction} times. The former is
  used to identify when the data was collected, whereas the latter when the data was
  transmitted to the data sink where the query processing acts. Finally, the
  \textbf{Where: Data Location} identifies the point where the collected data
  was observed by the sensor device. Usually the latitude and longitude as the
  GPS coordinates;
  \item \textbf{Annotations}: the annotation of the collected data, through the
  use of keywords, full-text description, or any other method that helps
  describing the collected data. It is important to note that this type of data
  does not make part of the collected data from sensors. Rather, they
  complement the observations of what was percieved by the sensor devices.
\end{itemize}

@@

\begin{figure}[h]
  \centering
  \includegraphics{../diagrams/taxonomy-query-mechanism}
  \caption{The Query Processing Mechanism Taxonomy}
  \label{fig:taxonomy-query-mechanism}
\end{figure}

@@

\section{Data Volume Taxonomy}

When it comes to the data volume produced during a time frame, different ranges
were reported from different types of sensor networks. Depending on the types
of data produced such as numerical or images, they can be in the orders of
Megabytes (1MB = 1000KB = ~10\superscript{3} bytes), Gigabytes (1GB =
1000\superscript{3} = 10\superscript{9} bytes) and even Terabytes (1TB =
1000\superscript{4} = 10\superscript{12}) to future Petabytes (1PB =
1000\superscript{5} = 10\superscript{15} bytes) as shown in section
\ref{sec:data-load}. As a comparison of what is processed, Google reports to
process 20 Petabytes of data per day in their data centers
\cite{map-reduce-load}. In this way, the following taxa are related to the
different sizes of data produced by a sensor network, summarized in Figure
\ref{fig:taxonomy-data-volume}:

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.5]{../diagrams/taxonomy-data-volume}
  \caption{The Data Volume Taxonomy}
  \label{fig:taxonomy-data-volume}
\end{figure}

\begin{itemize}
  \item \textbf{Small}: sensor networks whose yearly produced data is in the
  range of MBs or to 1GB of data, usually stored in flat files directly in the
  operating system. They are usually collected from sensor devices that only
  generate numerical data;
  \item \textbf{Medium}: amount of data that ranges from GBs to 1TB data. In
  order to manage such amount of data using disk arrays;
  \item \textbf{Large}: when the amount of data produced by a sensor device
  reaches amounts of data compatible to data centers, in the orders ranging
  from TBs to 1PT of data. The construction of a data center specialized for
  data produced by sensor devices, taking into account the addition of data
  provenance, among others.
\end{itemize}

\section{System Organization Taxonomy}

Different approaches of are developed computational systems to deal the data
produced from sensor networks. As  
In this way, Figure \ref{fig:taxonomy-database-architecture} describes the
taxonomy of the database system organization as follows:

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.5]{../diagrams/taxonomy-database-architecture}
  \caption{The System Organization Taxonomy}
  \label{fig:taxonomy-database-architecture}
\end{figure}

\begin{itemize}
  \item \textbf{Centralized System}: a database system that runs in one single
  centralized host, being the focal purpose managing the collected data
  \cite{sn-intro01};
  \item \textbf{Distributed System}: a database system that can be composed by
  more than one node. Usually, the use of the pattern of Master-Slave
  mechanism or data ``Replication" is used to aliviate ``hot spots'' during
  concurrent processes of data collection from the sensor devices and data
  access from the final users. On the other hand, unique mechanisms can be
  used to disagragate the produced data from a concentrated one: the use 
  of database partioning, as related to the data-centric storage, where  
  each partition is responsible to hand data produced by a given type of 
  sensor, for instance.
\end{itemize}

@@

On the whole, the taxonomies described in this chapter were based solely on the
literature review of described in the previous chapter. They can be helpful in 
understanding of requirements of a data persistence layer for sensor networks
with different infrastructures. They are summarized on table
\ref{tab:taxonomies-list-summary}, and briefly related highlight these
relationships in order to better analyze a problem related to data persistence.

\begin{table}[!b]
    \label{tab:taxonomies-list-summary}
    \begin{center}
        \begin{tabular}{|p{170pt}|p{250pt}|}\hline 
        \textbf{Taxonomy Name} & \textbf{Associated Taxas}\\\hline 
        Purpose of the Sensor Data & Real-time Data Stream, Data Archival \\\hline 
        The Location of the Sensor Data & Local Storage, External Storage\\\hline
        Data Model & Schema-Dependent, Schema-less Models\\\hline 
        Data Description & Provenance-Aware, Annotations\\\hline
        Query Processing & In-Network Query, Centralized Query\\\hline 
        Data Volume & Small, Medium, Large\\\hline
        System Organization & Centralized System, Distributed System\\\hline
        \end{tabular}
    \end{center}
    \caption{Summary of the Data Persistence Taxonomies}
\end{table}

The taxonomy related to the purpose of the sensor data can be directly related
to the taxonomy of the location of the sensor data. While the taxon of the 
real-time data stream relates to the local data storage, the taxon related to
data archival correlates the storage of the collected data with the an external
storage one. In addition to that, the converse is also true.

The taxonomy related to the data model can be related to any taxon from the
taxon described from the system organization, to data description and query
processing and location of the collected data. In the former, the data of a
relational model is stored using SQL messages.

Similarly, the taxonomy related to the volume of data produced by a sensor
network usually determines the system organization used to handle and process
the collected data. When the volume is small, the use a single server is
required, usually using an external storage device, which relates to the
taxonomy for storage system. However, when the volume is medium to large, the
use of distributed systems is required to cope with high volume of data
processing. Techniques described for data replication is used to decrease the
load in hot spots of the architecture. On the other hand, approaches to
separate the data being written and read can be used such as Load Balancing or
Database Partition. As a consequence, the use of a Data-Centric storage
approach is used.
